Fluent statistical computing interfaces for biological data analysis

### 2020-07-15 Rmd/01-intro.Rmd
spelling

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. It is organised in 
according to the aforementioned workflow steps. To begin we discuss wrangling
biological data using a coherent represntation and programming interface; 
section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). Next, we look at integrating data and 
model outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
introduces a framework for incorporating and discovering 'interesting' genomic
regions, described further in chapters \@ref(ch:fluentGenomics) and 
\@ref(ch:intron). Third, we consider the challenges in visualising  
high dimensional data; section \@ref(sec:va) introduces a visual analytics 
approach for understanding non-linear dimension reduction techniques described
in chapter \@ref(ch:tsne). Lastly, in chapter 
\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
developments.

@@

Chapter \@ref(ch:tsne) introduces
a software package called *liminal* for constructing these views and
a user interaction framework for identifying distortions. We present several
case studies using data sets from both single-cell transciptomics and 
machine learning  for using our approach to diagnose the quality of 
results obtained via popular non-linear DR methods
like t-distributed stochastic neighbour embeddings (t-SNE).

>>>>

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. It is organised in 
according to the aforementioned workflow steps. To begin we discuss wrangling
biological data using a coherent representation and programming interface; 
section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). Next, we look at integrating data and 
model outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
introduces a framework for incorporating and discovering 'interesting' genomic
regions, described further in chapters \@ref(ch:fluentGenomics) and 
\@ref(ch:intron). Third, we consider the challenges in visualising  
high dimensional data; section \@ref(sec:va) introduces a visual analytics 
approach for understanding non-linear dimension reduction techniques described
in chapter \@ref(ch:tsne). Lastly, in chapter 
\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
developments.

@@

Chapter \@ref(ch:tsne) introduces
a software package called *liminal* for constructing these views and
a user interaction framework for identifying distortions. We present several
case studies using data sets from both single-cell transcriptomics and 
machine learning  for using our approach to diagnose the quality of 
results obtained via popular non-linear DR methods
like t-distributed stochastic neighbour embeddings (t-SNE).

### 2020-07-15 Rmd/06-conclusion.Rmd
spelling

Chapter \@ref(ch:plyranges), "plyranges: a grammar of genomic data
transformation" developed a new domain specific language for genomics data analysis.
Chapter \@ref(ch:fluentGenomics), "Fluent genomics with plyranges and tximeta"
described techniques integrating data along the genome, and emphasised
the importance of interoperability between analysis tools. Chapter 
\@ref(ch:intron), "Exploratory coverage analysis with superintronic and 
plyranges", tackled data  integration from a different angle by looking at 
multiple summaries of variables measured along the genome to putative regions of 
intron retention.  Chapter \@ref(ch:tsne), "Casting multiple shadows: 
high-dimensional  interactive data visualisation with tours and embeddings", 
explored pragmatic approaches to high dimensional data visualisation in light of
the rise of popular non-linear embedding methods. 
To finish, I will dicuss the implications of this work and provide 
suggestions for further research.

@@

The early chapters of this thesis explored the *tidy data* semantic and its
extension to range based genomics data. This culminated in the development
of the **plyranges** package and consequent explorations of the applicability
of its interface and use of the *tidy data* concept in chapters 
\@ref(ch:fluentGenomics) and \@ref(ch:intron). The package has been relatively 
successful: it has been downloaded `r plyranges_stats$no_downloads` times from 
`r plyranges_stats$no_ip` unique IP addresses since its availability on 
Bioconductor. I have also had the privilege of teaching workshops on **plyranges**
at Bioconductor conferences which also led to the development of the work in 
chapter\@ref(ch:fluentGenomics) and refined the development of the API. A 
broader impact of the work, has been the discussions around the
concepts of fluent interfaces and tidy data within the Bioconductor community,
which has led to several developments currently in place that are exploring
different approaches for fluent interfaces for other types of 'omics data.

@@

A limitation of the grammar as we have implemented it in 
**plyranges** is lack of scalability and computational speed for data sets
that do not fit in memory. We attempted several techniques for perfoming 
delayed operations over range-based data, however a more general approach
that allows for data stored on the cloud or in scientific data formats
like HDF5 that leverages existing Bioconductor frameworks would be useful.
We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an a
nalyst is able to do some very complex data transformations and resampling 
procedures via casting results into *GRanges* object. However, it is unclear 
whether the semantics of our grammar can be extended to data that can not be 
easily reshaped into long form tidy representations. Moreover, further work
is required to explore the design space of grammars for data 
transformations and grammars for graphics when the data large, multifaceted
and non-rectangular.

>>>>

Chapter \@ref(ch:plyranges), "plyranges: a grammar of genomic data
transformation" developed a new domain specific language for genomics data analysis.
Chapter \@ref(ch:fluentGenomics), "Fluent genomics with plyranges and tximeta"
described techniques integrating data along the genome, and emphasised
the importance of interoperability between analysis tools. Chapter 
\@ref(ch:intron), "Exploratory coverage analysis with superintronic and 
plyranges", tackled data  integration from a different angle by looking at 
multiple summaries of variables measured along the genome to putative regions of 
intron retention.  Chapter \@ref(ch:tsne), "Casting multiple shadows: 
high-dimensional  interactive data visualisation with tours and embeddings", 
explored pragmatic approaches to high dimensional data visualisation in light of
the rise of popular non-linear embedding methods. 
To finish, I will discuss the implications of this work and provide 
suggestions for further research.

@@

The early chapters of this thesis explored the *tidy data* semantic and its
extension to range based genomics data. This culminated in the development
of the **plyranges** package and consequent explorations of the applicability
of its interface and use of the *tidy data* concept in chapters 
\@ref(ch:fluentGenomics) and \@ref(ch:intron). The package has been relatively 
successful: it has been downloaded `r plyranges_stats$no_downloads` times from 
`r plyranges_stats$no_ip` unique IP addresses since its availability on 
Bioconductor. I have also had the privilege of teaching workshops on **plyranges**
at Bioconductor conferences which also led to the development of the work in 
chapter\@ref(ch:fluentGenomics) and refined the development of the API. A 
broader impact of the work, has been the discussions around the
concepts of fluent interfaces and tidy data within the Bioconductor community,
which has led to several developments currently in place that are exploring
different approaches for fluent interfaces for other types of omics data.

@@

A limitation of the grammar as we have implemented it in 
**plyranges** is lack of scalability and computational speed for data sets
that do not fit in memory. We attempted several techniques for performing 
delayed operations over range-based data, however a more general approach
that allows for data stored on the cloud or in scientific data formats
like HDF5 that leverage existing Bioconductor frameworks would be useful.
We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an a
analyst is able to do some very complex data transformations and re-sampling 
procedures via casting results into *GRanges* object. However, it is unclear 
whether the semantics of our grammar can be extended to data that can not be 
easily reshaped into long form tidy representations. Moreover, further work
is required to explore the design space of grammars for data 
transformations and grammars for graphics when the data large, multifaceted
and non-rectangular.
